A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-10-08 paper/paper.tex
paper: lots of minor corrections

Unless otherwise specified, we fix a universe level $m$ and abbreviate
$\Type_m$ as $\Type$ and $\U[m]$ as $\U$.

We are also interested in \emph{extensional} type theory; i.e. in a type theory with the following reflection rule, where $=$ is propositional equality in some type and $\equiv$ is definitional equality~\cite{martinlof84}:
\begin{equation*}
  \seqr{}{\Gamma \vdash p: t = u}{\Gamma \vdash t \equiv u}
\end{equation*}

@@

We use the type $\U$ to represent $\nu$-sets in type theory and fix for that purpose a universe level $m$, that is, we define $\nu$-sets at level $m$ living in $\U[m+1]$.

Table \ref{tab:coind} describes a $\nu$-set at level $m$ in indexed form, as a coinductively-defined infinite sequence of types representing the limit of $n$-truncated $\nu$-sets: $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n+1}, X_{n+2}, \ldots$ dependent on a $n$-truncated $\nu$-set $\Xto{m}{n}$ so that, when $m$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $0$-truncated $\nu$-set $\Xto{m}{n}$ is degenerated: it is defined to be the unit type and there is thus only one $0$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

The definition of a $n$-truncated $\nu$-set is in turn described in \ref{tab:core}. In the infinite sequence of types representing a $\nu$-set, the $n$-th component is a type dependent over a $\fullframe$. The type $\fullframe$ represents a matching object. It is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\framep$, $\layer$ and $\painting$. The rationale behind $\framep$, $\layer$ and $\painting$ will be explained later but it should already be noticed that the type $\layer$ relies on an operator of frame restriction $\restrf$ which is defined in \ref{tab:faces}, and this restriction operator is in turn defined using the auxiliary definitions of $\restrl$ and $\restrc$.

@@

\subsection*{Intuition for our formal construction\label{sec:intuition}}
\renewcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\renewcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\renewcommandx{\painting}[2][1,2]{\prim{painting}[][#1][#2][][]}
\renewcommandx{\restrf}[4][1,2,3,4]{\restr{frame}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrl}[4][1,2,3,4]{\restr{layer}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrc}[4][1,2,3,4]{\restr{painting}[][#3][#4][#1][#2][][]}
\renewcommandx{\cohf}[1][1]{\coh{frame}[][][][][][#1][][]}
\renewcommandx{\cohl}[1][1]{\coh{layer}[][][][][][#1][][]}
\renewcommandx{\cohc}[1][1]{\coh{painting}[][][][][][#1][][]}
\renewcommandx{\coht}{\cohtwo{frame}[][][][][][][][]}

To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on building blocks, which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$.

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying to the description above the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A.\, (B a \rightarrow C)$ and $(\Sigma a: A.\, B a) \rightarrow C$. In particular, $\fullframe[n]$ is a ``telescope'', i.e. a nesting of $\Sigma$-type. Let us now illustrate how we build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown below. The figure also mentions how the type of $X_2$ is structured.

@@

The figure on the left is $\framep[3][1]$, on the right is $\framep[3][2]$, and on the bottom is $\framep[3][3]$, which is full. Further, $\framep[3][1]$ is made of one layer, $\layer[3][0]$, shown in blue, $\framep[3][2]$ is made of one additional layer, $\layer[3][1]$, shown in red, $\framep[3][3]$ is made of one more layer, $\layer[3][2]$, shown in green.

@@

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a nesting of $\Sigma$-types, which we abbreviate as $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.
\begin{equation*}
  \begin{array}{lll}
    \framep[n][p](D)   & \defeq & \Sigma a_n: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)) \ldots) .\,\layer[n][p - 1](D) \\
    \painting[n][p](D) & \defeq & \Sigma l_p: \layer[n][p](D).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D) .\, D.\tl) \ldots)
  \end{array}
\end{equation*}

An extra refinement arises from the fact that each addition of a layer to a frame has to be glued onto the border of the partial frame built so far. So, each $\layer[n][p]$ has to depend on $\framep[n][p]$. We also need a way to characterize the border of the $\nu$ $\painting[n-1][p]$ that compose a $\layer[n][p]$, and this is $\restrf[n][p][\epsilon][p]$, for all $\epsilon < \nu$. For instance, on the picture, the left and right $\painting[2][1]$, shown in red, are laid on respectively the left and right borders of the blue boxes, and hence needs to depend on $\framep[3][1]$. The left and right borders of the two blue boxes are extracted as $\restrf[2][1][L](D)(d)$ and $\restrf[2][1][R](D)(d)$. We can then refine the previous equation by showing the dependencies on $d$, as shown below.

@@

The operation $\restrf[n][p][\epsilon][q]$ indicates restriction on $q$ layers of a frame, and the induction is on $p$, from $0$ to $q$. In particular, $\restrf[n][p][\epsilon][p]$ is a ``full restriction''. We define $\restrf[n][p][\epsilon][p](d)$ by recursion on the structure of a frame $d$, which necessitates definitions of $\restrl[n][p][\epsilon][q](d)(l)$ and $\restrc[n][p][\epsilon][q](d)(c)$, for $l$ a $\layer$ and $c$ a $\painting$. The key case is $\restrc[n][p][\epsilon][p](d)(c)$, where $c$, a $\painting[n][p]$, has necessarily the form of $((c_L, c_R), \_)$: $\restrc[n][p][L][p]$ picks out $c_L$, a $\painting[n-1][p]$, $\restrc[n][p][R][p]$ picks out the $c_R$, also a $\painting[n-1][p]$, and $\_$, a $\painting[n][p+1]$, is discarded. There is one more difficulty, which we illustrate by writing down expected and actual types.

@@

\subsection{Groupoid properties of equality and basic type isomorphisms\label{sec:eqproperties}}

The construction takes benefit of various provable equalities over proofs of equality being definitional by the reflection rule. This includes in particular the groupoid properties of equality. Notably, uniqueness of identity proofs holds in extensional type theory, so that any type is automatically an $\U$. Also, we left implicit in table~\ref{tab:coh} the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type. In the same table, we also left implicit the use of the isomorphism between $f = g$ and $\Pi a: A.\, f(a) = g(a)$ for $f$ and $g$ in $\Pi a: A.\, B$, where it should be recalled that the right-to-left map, that is function extensionality, holds by default in extensional type theory.

>>>>

By the notations $\Type$ and $\U$ will be meant $\Type_m$ or $\U[m]$ at some unspecified universe level $m$.

We are also interested in \emph{extensional} type theory; i.e. in a type theory with the following reflection rule, where $=$ is propositional equality in some type and $\equiv$ is definitional equality~\citep{martinlof84}:
\begin{equation*}
  \seqr{}{\Gamma \vdash p: t = u}{\Gamma \vdash t \equiv u}
\end{equation*}

@@

A $\nu$-set is a sequence of families of $\U$, that is $\U[m]$ for some universe level $m$. We call such sequence a $\nu$-set at level $m$, whose type thus lives in $\U[m+1]$.

Table \ref{tab:coind} describes the type of a $\nu$-set at level $m$ in indexed form, as the type of a coinductively-defined infinite sequence of type families representing the limit of $n$-truncated $\nu$-sets: $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n+1}, X_{n+2}, \ldots$ dependent on a $n$-truncated $\nu$-set, $\Xto{m}{n}$, so that, when $m$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $0$-truncated $\nu$-set, $\Xto{m}{0}$, is degenerated: it is an empty family, defined to be the unit type, and there is thus only one $0$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

The definition of the type of a $n$-truncated $\nu$-set is in turn described in \ref{tab:core}. In the infinite sequence of type families representing a $\nu$-set, the $n$-th component is a type dependent over a $\fullframe$. The type $\fullframe$ represents a matching object. It is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\framep$, $\layer$ and $\painting$. The rationale behind $\framep$, $\layer$ and $\painting$ will be explained later but it should already be noticed that the type $\layer$ relies on an operator of frame restriction $\restrf$ which is defined in \ref{tab:faces}, and this restriction operator is in turn defined using the auxiliary definitions of $\restrl$ and $\restrc$.

@@

\subsection*{Intuition for our formal construction\label{sec:intuition}}
\renewcommandx{\fullframe}[1][1]{\prim{fullframe}[][#1][][][]}
\renewcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\renewcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\renewcommandx{\painting}[2][1,2]{\prim{painting}[][#1][#2][][]}
\renewcommandx{\restrf}[4][1,2,3,4]{\restr{frame}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrl}[4][1,2,3,4]{\restr{layer}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrc}[4][1,2,3,4]{\restr{painting}[][#3][#4][#1][#2][][]}
\renewcommandx{\cohf}[1][1]{\coh{frame}[][][][][][#1][][]}
\renewcommandx{\cohl}[1][1]{\coh{layer}[][][][][][#1][][]}
\renewcommandx{\cohc}[1][1]{\coh{painting}[][][][][][#1][][]}
\renewcommandx{\coht}{\cohtwo{frame}[][][][][][][][]}

To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on the building blocks which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$: they represent matching objects.

To every $X_n$ is uniformly assigned a type of the form $\fullframe[n] \rightarrow \U$. Compared to the informal description given in section~\ref{sec:rel-param}, e.g. $X_1 : X_0 \rightarrow X_0 \rightarrow \U$, this amounts to apply the isomorphisms between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A.\, (B a \rightarrow C)$ and $(\Sigma a: A.\, B a) \rightarrow C$ so as to get the expected form $\fullframe[n] \rightarrow \U$.. In particular, $\fullframe[n]$ is a ``telescope'', i.e. a nesting of $\Sigma$-types.

Let us now illustrate how we recursively build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ given to $X_0$ in section~\ref{sec:rel-param} can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown below. The figure also mentions how the type of $X_2$ is structured.

@@

The figure on the left is $\framep[3][1]$, in the middle is $\framep[3][2]$, and on the right is $\framep[3][3]$, which is full. Further, $\framep[3][1]$ is made of one layer, $\layer[3][0]$, shown in blue, $\framep[3][2]$ is made of one additional layer, $\layer[3][1]$, shown in red, $\framep[3][3]$ is made of one more layer, $\layer[3][2]$, shown in green.

@@

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$-type. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a nesting of $\Sigma$-types, constituting the type of a $n$-truncated $\nu$-set, which we wrote $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then reformulate the previous equation with its dependency on $D$. In particular, $X_n$ is just the last component of $D$, that is $D.\tl$.
\begin{equation*}
  \begin{array}{lll}
    \framep[n][p](D)   & \defeq & \Sigma a_n: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)) \ldots) .\,\layer[n][p - 1](D) \\
    \painting[n][p](D) & \defeq & \Sigma l_p: \layer[n][p](D).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D) .\, D.\tl) \ldots)
  \end{array}
\end{equation*}

An extra refinement arises from the fact that each new layer of a frame has to be glued onto the border of the partial frame built so far. So, each $\layer[n][p]$ has to depend on $\framep[n][p]$. We also need a way to characterize the border of the $\nu$ $\painting[n-1][p]$ that compose a $\layer[n][p]$, and this is where the restriction $\restrf[n][p][\epsilon][p]$ is involved, for each $\epsilon < \nu$. For instance, on the picture, the left and right $\painting[2][1]$, shown in red, are laid on respectively the left and right borders of the blue boxes, and hence needs to depend on $\framep[3][1]$. The left and right borders of the two blue boxes are extracted as $\restrf[2][1][L](D)(d)$ and $\restrf[2][1][R](D)(d)$. We can then refine again the previous equation by showing the dependencies on $d$, as shown below.

@@

The operation $\restrf[n][p][\epsilon][q]$ restricts the $p$ first layers of a frame, and the induction is on $p$, from $0$ to $q$. In particular, $\restrf[n][p][\epsilon][p]$ is a ``full restriction''. We define $\restrf[n][p][\epsilon][q](d)$ by recursion on the structure of a frame $d$, which necessitates definitions of $\restrl[n][p][\epsilon][q](d)(l)$ and $\restrc[n][p][\epsilon][q](d)(c)$, for $l$ a $\layer$ and $c$ a $\painting$. The key case is $\restrc[n][p][\epsilon][p](d)(c)$, where $c$, a $\painting[n][p]$, has necessarily the form of $((c_L, c_R), \_)$: $\restrc[n][p][L][p]$ picks out $c_L$, a $\painting[n-1][p]$, $\restrc[n][p][R][p]$ picks out the $c_R$, also a $\painting[n-1][p]$, and $\_$, a $\painting[n][p+1]$, is discarded. There is one more difficulty, which we illustrate by writing down expected and actual types.

@@

\subsection{Groupoid properties of equality and basic type isomorphisms\label{sec:eqproperties}}
The construction takes benefit of various provable equalities over proofs of equality being definitional by the reflection rule. This includes in particular the groupoid properties of equality. Notably, uniqueness of identity proofs holds in extensional type theory, so that any type is automatically an $\U$. Also, we left implicit in table~\ref{tab:coh} the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type. In the same table, we also left implicit the use of the isomorphism between $f = g$ and $\Pi a: A.\, f(a) = g(a)$ for $f$ and $g$ in $\Pi a: A.\, B$, where it should be recalled that the right-to-left map, that is function extensionality, holds by default in extensional type theory.
